---
title: "PancCanNet Workflow 3 PPI network analysis"
author: "mkutmon"
date: "1 September 2021"
version: 1.0
license: "MIT License"
output: 
  md_document:
    variant: markdown_github
always_allow_html: true
---

# General instructions (read before running the code snippets)
In this second workflow, we will create a protein-protein interaction network of the up- and down-regulated genes in the different pancreatic cancer subtypes. Afterwards, we will extend the network with gene-pathway associations to see in which pathways the differentially expressed genes are present in.

* The script contains several code snippets which should be run one after the other. 
* Make sure all the required packages are installed beforehand (BiocManager::install(...)). 
* Make sure you have Cytoscape installed (version 3.8.0+) and running before you start running the script. 

*** 

## R environment setup

First, we need to make sure all required R-packages are installed. Run the following code snippet to install and load the libraries. 
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if(!"dplyr" %in% installed.packages()) BiocManager::install("dplyr")
if(!"rWikiPathways" %in% installed.packages()) BiocManager::install("rWikiPathways")
if(!"RCy3" %in% installed.packages()) BiocManager::install("RCy3")
if(!"RColorBrewer" %in% installed.packages()) BiocManager::install("RColorBrewer")
if(!"rstudioapi" %in% installed.packages()) BiocManager::install("rstudioapi")
if(!"readr" %in% installed.packages()) BiocManager::install("readr")
if(!"org.Hs.eg.db" %in% installed.packages()) BiocManager::install("org.Hs.eg.db")


library(dplyr)
library(rWikiPathways)
library(RCy3)
library(RColorBrewer)
library(rstudioapi)
library(readr)
library(org.Hs.eg.db)

setwd(dirname(rstudioapi::getSourceEditorContext()$path))
```

*** 

## Load differential gene expression dataset

In the following section, you will load a differential gene expression dataset. You can easily replace the example dataset with your own dataset (copy it in the "data" folder and change the file name below). 
In this workflow, we will identify affected pathways in two pancreatic cancer subtypes, and visualize the data on the pathway models. 

```{r}
dataset <- read.delim("data/GSE71729-dataset.txt")

# filter genes without Entrez Gene identifier
data.panc <- dataset %>% tidyr::drop_na(Entrez.Gene)
colnames(data.panc)[2] <- "GeneName"
colnames(data.panc)[1] <- "GeneId"

# DEG in either classical or basal subtype
# stricter cutoff in this example
fc_cutoff <- 1

deg <- unique(data.panc[(data.panc$B_adj.P.Val < 0.05 & abs(data.panc$B_logFC) > fc_cutoff) | (data.panc$C_adj.P.Val < 0.05 & abs(data.panc$C_logFC) > fc_cutoff),c(1,2)])

```

***

# PPI network analysis

Next, we will create a protein-protein interaction network with all differentially expressed genes using the STRING database. 

```{r}
RCy3::cytoscapePing()
installApp('stringApp') 

query <- format_csv(as.data.frame(deg$GeneId), col_names=F, escape = "double", eol =",")
commandsRun(paste0('string protein query cutoff=0.9 newNetName="PPI network" query="',query,'" limit=0'))

# network will be opened in Cytoscape (this might take a while)
```

> Let's explore the network

- **Q1**: How many of the differentially expressed genes were found in STRING?
- **Q2**: Are all genes connected in the network? 
- **Q3**: Change the confidence cutoff in the commandsRun call from 0.9 (high confidence) to 0.4 (medium confidence). What changes? 

***

## Data visualization

Let's now visualize the gene expression data on the PPI network.

```{r}
loadTableData(data.panc, data.key.column = "GeneId", table.key.column = "query term")

RCy3::installApp("enhancedGraphics")
RCy3::copyVisualStyle("default", "my_style_heatmap")
RCy3::setNodeLabelMapping("display name", style.name = "my_style_heatmap")
RCy3::setNodeCustomHeatMapChart(c("B_logFC","C_logFC"), slot = 2, style.name = "my_style_heatmap", colors = c("#CC3300","#FFFFFF","#6699FF","#CCCCCC"))

RCy3::setVisualStyle("my_style_heatmap")
```

> Interpretation

- **Q4**: Do you see clusters of up- or down-regulated genes in the PPI network?

***

## Pathway information

Next, we will add information about participation of the differentially expressed genes in molecular pathway models.

```{r}
# run CyTargetLinker
RCy3::installApp("CyTargetLinker")

wp <- file.path(getwd(), "data/wikipathways-hsa-20200710.xgmml")

commandsRun(paste0('cytargetlinker extend idAttribute="query term" linkSetFiles="', wp, '"'))
commandsRun('cytargetlinker applyLayout network="current"')
commandsRun('cytargetlinker applyVisualstyle network="current"')
RCy3::setNodeLabelMapping("display name", style.name="CyTargetLinker")

# there is an issue in the latest version with visualization of the added edges - the workaround below solves this for now
suid.ctl <- RCy3::cloneNetwork()
RCy3::setCurrentNetwork(suid.ctl)
RCy3::setVisualStyle("default")
RCy3::setVisualStyle("CyTargetLinker")

# TODO: VISUAL STYLE
loadTableData(data.panc, data.key.column = "GeneId", table.key.column = "query term")
RCy3::setNodeCustomHeatMapChart(c("B_logFC","C_logFC"), slot = 2, style.name = "CyTargetLinker", colors = c("#CC3300","#FFFFFF","#6699FF","#CCCCCC"))

png.file <- file.path(getwd(), "output/w3-fig1.png")
exportImage(png.file,'PNG', zoom = 500)

```

> Interpretation

- **Q5**: How many differentially expressed genes are in at least one of the pathways? 
- **Q6**: Are the genes also functionally related based on the PPI network?

***

## Save Cytoscape output and session

```{r}
# Saving output

cys.file <- file.path(getwd(), "output/workflow3.cys")
saveSession(cys.file) 

#comment following line if you want to manipulate the visualization in Cytoscape
RCy3::closeSession(save.before.closing = F)
```